Type-safe client-server communication for web applications

### 2014-02-28 paper.tex
Explain App monad.

@@

Functions are represented as plain Javascript functions, one of the blessings
of targeting a high level language, and application is can therefore be
implemented as its Javascript counterpart in most cases. In the general case,
however, functions may be curried. For such cases where the arity of an applied
function can not be determined statically, application is implemented using the
eval/apply method described in \cite{fastcurry} instead.

@@

This approach was inspired by Cloud Haskell \cite{cloudhaskell}, which
introduces the notion of ``static values''; values which are known at compile
time. Codifying this in the type system, to enable it to be used as a basis for
remote procedure calls, unfortunately requires some major changes to the
compiler. Cloud Haskell has a stopgap measure for unmodified compilers wherein
a remote table, pairing values with unique identifiers, are kept. This explicit
bookkeeping relies on the programmer to assign appropriate types to both values
themselves and their identifiers, breaking type safety.

@@

Another approach to this problem is defunctionalization
\cite{defunctionalization}, a program transformation wherein functions are
translated into algebraic data types. This approach would allow the client and
server to use the same actual code; rather than passing an identifier around,
the client would instead pass the actual defunctionalized function. This would
have the added benefit of allowing functions to be arbitrarily composed before
being remotely invoked.

This approach also requires significant changes to the compiler, making it
unsuitable for our use case. Moreover, we are not entirely convinced about the
wisdom of allowing server-side execution of what is essentially arbitrary code
sent from the client which, for web applications, is completely untrustworthy.

@@

\acks

Acknowledgments, if needed.

>>>>

Our implementation is built in three layers: the compiler layer, the
concurrency layer and the communication layer. The concurrency and
communication layers are simple Haskell libraries, portable to any other pair
of standard Haskell compilers with minimal effort.

\paragraph{Two compilers}
The principal trick to our solution is compiling the same program twice; once
with a compiler that generates the server binary, and once with one that
generates Javascript for execution on the client. Conditional compilation is
used for a select few functions, to enable slightly different behavior on the
client and on the server as necessary. As we believe that a monadic
programming style is a good fit for this application, we have chosen to use
Haskell as the base of our solution, which by default leads us to choose GHC
as our server-side compiler. For reasons outlined in section \ref{sec:haste},
the Haste compiler was chosen to provide the client-side code.

\paragraph{The \lstinline!App! monad}
The \lstinline!App! monad is where remote functions are declared, server state
is initialized and program flow is finally handed over to the \lstinline!Client!
monad. Slightly simplified, its definition is as follows.

\begin{code}
type Method = [JSON] -> IO JSON
type AppState = (CallID, [(CallID, Method)])
newtype App a = App (StateT AppState IO a)
\end{code}

As we can see, \lstinline!App! is a simple state monad, with underlying IO
capabilities to allow server-side computations to be forked from within it.
Aside from the \lstinline!runClient! function, which is only used to launch the
client program, This monad has two primary operations.

\begin{code}
liftServerIO :: IO a -> App (Useless a)
export :: Exportable a => a -> App (Export a)
\end{code}

Aside from the \lstinline!App! monad itself, these functions operate on two
types we have not seen before.

\begin{code}
data Useless a = Useful a | Useless
data Export a = Export CallID [JSON]
\end{code}

The definition of \lstinline!liftServerIO! is very simple. Through the use of
conditional compilation, on the client it simply ignores its argument and
returns \lstinline!Useless!; recall that server-lifted IO computations are not
observable on the client. On the server side, \lstinline!liftServerIO! executes
its argument and wraps the computation's return value using the
\lstinline!Useful! constructor. The counterpart function, \lstinline!mkUseful!
simply unwraps the \lstinline!Useful! constructor and returns the contained
value in the \lstinline!Server! monad.

\paragraph{The \lstinline!Client! monad and the \lstinline!onServer! function}
As synchronous network communication is one of our stated goals, it is clear
that we will need some kind of blocking primitive. Since Javascript does not
support any kind of blocking, we will have to implement this ourselves.

A solution is given in the \emph{poor man's concurrency
monad}\ \cite{concurrencymonad}. Making use of a continuation monad with
primitive operations for forking a computation and atomically lifting an IO
computation into the monad, it is possible to implement cooperative
multitasking on top of the non-concurrent Javascript runtime. This monad
allows us to implement \lstinline!MVar!s as our blocking primitive, with the
same semantics as their regular Haskell counterpart.\ \cite{ffi}
This concurrency-enhanced monad is used as the basis of the \lstinline!Client!
monad.

@@

Functions are represented as plain Javascript functions, one of the blessings
of targeting a high level language, and application is can therefore be
implemented as its Javascript counterpart in most cases. In the general case,
however, functions may be curried. For such cases where the arity of an applied
function can not be determined statically, application is implemented using the
eval/apply method described in\ \cite{fastcurry} instead.

@@

This approach was inspired by Cloud Haskell\ \cite{cloudhaskell}, which
introduces the notion of ``static values''; values which are known at compile
time. Codifying this in the type system, to enable it to be used as a basis for
remote procedure calls, unfortunately requires some major changes to the
compiler. Cloud Haskell has a stopgap measure for unmodified compilers wherein
a remote table, pairing values with unique identifiers, are kept. This explicit
bookkeeping relies on the programmer to assign appropriate types to both values
themselves and their identifiers, breaking type safety.

@@

Another approach to this problem is
defunctionalization\ \cite{defunctionalization}, a program transformation
wherein functions are translated into algebraic data types. This approach would
allow the client and server to use the same actual code; rather than passing
an identifier around, the client would instead pass the actual defunctionalized
function. This would have the added benefit of allowing functions to be
arbitrarily composed before being remotely invoked.

This approach also requires significant changes to the compiler, making it
unsuitable for our use case. Moreover, we are not entirely convinced about the
wisdom of allowing server-side execution of what is essentially arbitrary code
sent from the client which, for web applications, is completely untrustworthy.
While analyzing code for improper behavior is certainly possible, implementing
and enforcing a sufficiently strict-yet-flexible security policy for remote
execution would be an unwelcome burden on the programmer.

@@

%\acks
%
%Acknowledgments, if needed.
